---
title: "STAT PROJECT"
author: "Jyot Patel"
date: "2026-02-27"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE)


```

```{r}
# Verify working directory and data availability
getwd()
list.files("data/lgg_tcga_pan_can_atlas_2018")
```

`############################################################

                    1. Load Clinical Data

`############################################################

```{r task3-load-clinical-patient, message=FALSE, warning=FALSE}


#####################  Load Clinical Patient   #####################################



# Load the data while skipping the first 4 metadata rows
clinical_patient <- read.delim(
  "data/lgg_tcga_pan_can_atlas_2018/data_clinical_patient.txt",
  sep = "\t",
  header = TRUE,
  skip = 4,
  stringsAsFactors = FALSE
)

# Display the first 10 rows
head(clinical_patient, 10)


```

```{r}

#########################  Load Clinical Sample #################################
clinical_sample <- read.delim(
  "data/lgg_tcga_pan_can_atlas_2018/data_clinical_sample.txt",
  sep = "\t",
  header = TRUE,
  skip = 4,
  stringsAsFactors = FALSE
)

head(clinical_sample, 10)
```

```{r}

# connection between the PATIENT_IDs contained in the patient clinical data and the SAMPLE_IDs contained in the sample clinical data

#Each SAMPLE_ID belongs to exactly one PATIENT_ID.
#But a PATIENT_ID can have:
#One sample OR multiple samples
#So the relationship is: One-to-many: one patient → multiple samples

table(clinical_sample$PATIENT_ID)

#If some counts are > 1 → that patient has multiple samples.

length(unique(clinical_patient$PATIENT_ID))
length(unique(clinical_sample$PATIENT_ID))

```

`############################################################

                2. Load mRNA Expression Data

`############################################################

```{r}

############################## Load mRNA #########################################

mrna_data <- read.delim(
  "data/lgg_tcga_pan_can_atlas_2018/data_mrna_seq_v2_rsem.txt",
  sep = "\t",
  header = TRUE,
  #nrows = 250,              # ← only first 250 genes
  stringsAsFactors = FALSE,
  check.names = FALSE
)

head(mrna_data, 10)

dim(mrna_data)

num_genes <- nrow(mrna_data)
num_samples <- ncol(mrna_data) - 1

mrna_raw <- mrna_data

```

`############################################################

                     3. Data Cleaning

`############################################################

```{r}
# --1) Drop irrelevant variables (metadata / redundant IDs)
# Entrez_Gene_Id is redundant for modeling here (gene name is enough).
mrna_clean <- mrna_raw

# Check if Entrez_Gene_Id exists
has_entrez <- "Entrez_Gene_Id" %in% colnames(mrna_raw)

if (has_entrez) {
  mrna_clean$Entrez_Gene_Id <- NULL
  cat("\n[CLEAN] Dropped Entrez_Gene_Id (redundant gene identifier).\n")
} else {
  cat("\n[NOTE] Entrez_Gene_Id not present — nothing to drop.\n")
}

cat("[CLEAN] Dimensions after dropping irrelevant columns:", 
    nrow(mrna_clean), ncol(mrna_clean), "\n")
```

```{r}

##### How many missing rows in Hugo_symbol ###########

sum(is.na(mrna_raw$Hugo_Symbol) | mrna_raw$Hugo_Symbol == "")

```

```{r}

# --- Drop genes with missing Hugo_Symbol (standardized gene name)
# Count missing Hugo_Symbol
missing_hugo <- sum(is.na(mrna_clean$Hugo_Symbol) | mrna_clean$Hugo_Symbol == "")
cat("\n[CHECK] Missing Hugo_Symbol rows:", missing_hugo, "\n")

# Remove those rows
before <- nrow(mrna_clean)
mrna_clean <- mrna_clean[!(is.na(mrna_clean$Hugo_Symbol) | mrna_clean$Hugo_Symbol == ""), ]
after <- nrow(mrna_clean)

cat("[CLEAN] Dropped rows with missing Hugo_Symbol.\n")
cat("        Rows before:", before, "Rows after:", after, "Removed:", before - after, "\n")

```

```{r}

# --2) Check data types
# Hugo_Symbol should be character; all sample columns should be numeric.
cat("\n[CHECK] Column classes (first 5 columns after cleaning):\n\n")
print(sapply(mrna_clean[, 1:min(5, ncol(mrna_clean))], class))

# Identify sample columns = everything except Hugo_Symbol
sample_cols <- setdiff(colnames(mrna_clean), "Hugo_Symbol")

# Check how many sample columns are numeric
is_num <- sapply(mrna_clean[, sample_cols, drop=FALSE], is.numeric)
cat("\n[CHECK] Numeric sample columns:", sum(is_num), "out of", length(sample_cols), "\n")

# If any sample columns are not numeric, coerce them safely
if (any(!is_num)) {
  bad_cols <- names(is_num)[!is_num]
  cat("[CLEAN] Non-numeric sample columns detected:", length(bad_cols), "\n")
  cat("        Coercing them to numeric (may introduce NAs if parsing fails).\n")
  
  for (cc in bad_cols) {
    mrna_clean[[cc]] <- suppressWarnings(as.numeric(mrna_clean[[cc]]))
  }
  
  # Re-check after coercion
  is_num2 <- sapply(mrna_clean[, sample_cols, drop=FALSE], is.numeric)
  cat("[CLEAN] Numeric sample columns after coercion:", sum(is_num2), "out of", length(sample_cols), "\n")
}
```

```{r}

# --3) Check data domains (are expression values valid?)
# For RSEM normalized expected read counts: values should be >= 0 (non-negative).
mat <- as.matrix(mrna_clean[, sample_cols, drop=FALSE])

min_val <- suppressWarnings(min(mat, na.rm = TRUE))
max_val <- suppressWarnings(max(mat, na.rm = TRUE))

cat("\n[CHECK] Expression domain:\n")
cat("  - Min value:", min_val, "\n")
cat("  - Max value:", max_val, "\n")

# Count negatives
neg_count <- sum(mat < 0, na.rm = TRUE)
cat("  - Negative values count:", neg_count, "\n")

# If negatives exist, you must decide what to do (usually they shouldn't exist)
# Here we DO NOT auto-fix them; we flag them loudly.
if (neg_count > 0) {
  cat("[WARNING] Negative expression values found. This is unexpected for RSEM counts.\n")
  cat("          You should investigate (file version, transformation, parsing).\n")
}

```

```{r}

# --4) Check missing data (NAs)
na_total <- sum(is.na(mrna_clean))
na_by_col <- sort(colSums(is.na(mrna_clean)), decreasing = TRUE)

cat("\n[CHECK] Missing data (NAs):\n")
cat("  - Total NA values in dataset:", na_total, "\n")
cat("  - Columns with NAs (top 10):\n")
print(head(na_by_col[na_by_col > 0], 10))

# If NAs exist, a common choice is to drop genes with ANY NA across samples
# Justification: gene expression should be complete; NAs complicate modeling.
if (na_total > 0) {
  before <- nrow(mrna_clean)
  mrna_clean <- mrna_clean[complete.cases(mrna_clean), ]
  after <- nrow(mrna_clean)
  cat("[CLEAN] Dropped rows (genes) with any NA.\n")
  cat("        Rows before:", before, "Rows after:", after, "Rows removed:", before - after, "\n")
  
  # Recompute sample matrix after filtering
  sample_cols <- setdiff(colnames(mrna_clean), "Hugo_Symbol")
  mat <- as.matrix(mrna_clean[, sample_cols, drop=FALSE])
  cat("        Total NAs after removal:", sum(is.na(mrna_clean)), "\n")
} else {
  cat("[CLEAN] No NAs found. No rows removed for missingness.\n")
}

```

```{r}

# --5) Check for duplicate genes (duplicate Hugo_Symbol)
dup_gene_count <- sum(duplicated(mrna_clean$Hugo_Symbol))
cat("\n[CHECK] Duplicate Hugo_Symbol count:", dup_gene_count, "\n")

# If duplicates exist, keep the first occurrence (simple, transparent rule).
# Alternative: average duplicates, keep max variance, etc. But keep-first is acceptable if justified.
if (dup_gene_count > 0) {
  before <- nrow(mrna_clean)
  mrna_clean <- mrna_clean[!duplicated(mrna_clean$Hugo_Symbol), ]
  after <- nrow(mrna_clean)
  cat("[CLEAN] Removed duplicate genes by keeping the first occurrence.\n")
  cat("        Rows before:", before, "Rows after:", after, "Removed:", before - after, "\n")
}

```

```{r}

# --6) Check for duplicate sample columns (duplicate SAMPLE IDs)
# In wide format, duplicated column names mean duplicated samples.
dup_sample_cols <- sum(duplicated(sample_cols))
cat("\n[CHECK] Duplicate sample columns (by column name):", dup_sample_cols, "\n")

# If duplicates exist, keep the first (again, simple rule).
if (dup_sample_cols > 0) {
  keep_cols <- !duplicated(colnames(mrna_clean))
  mrna_clean <- mrna_clean[, keep_cols]
  cat("[CLEAN] Removed duplicated sample columns by keeping first occurrence.\n")
  cat("        New dimensions:", nrow(mrna_clean), ncol(mrna_clean), "\n")
  sample_cols <- setdiff(colnames(mrna_clean), "Hugo_Symbol")
  mat <- as.matrix(mrna_clean[, sample_cols, drop=FALSE])
}

```

```{r}

# --- 8) Executive summary (print final stats)
cat("\n=========================\n")
cat("EXECUTIVE SUMMARY (mRNA CLEANING)\n")
cat("=========================\n")
cat("Final dimensions (rows=genes, cols=1 + samples):", nrow(mrna_clean), ncol(mrna_clean), "\n")
cat("Final genes:", nrow(mrna_clean), "\n")
cat("Final samples:", ncol(mrna_clean) - 1, "\n")
cat("Total NAs now:", sum(is.na(mrna_clean)), "\n")
cat("Duplicate genes now:", sum(duplicated(mrna_clean$Hugo_Symbol)), "\n")
cat("Min/Max expression now:", min(as.matrix(mrna_clean[, -1]), na.rm=TRUE), "/", 
                               max(as.matrix(mrna_clean[, -1]), na.rm=TRUE), "\n")

#  9) Display first 10 rows (as required)
cat("\n[OUTPUT] First 10 rows of cleaned mRNA data:\n")
head(mrna_clean, 10)

```

```{r}

# =========================
# Assignment 2 - Part B
# Quality check + clean patient clinical data
# =========================

# Keep only required columns
clinical_patient_clean <- clinical_patient[, c("PATIENT_ID", "SUBTYPE")]

cat("\n[CLEAN] Dimensions after keeping PATIENT_ID and SUBTYPE:",
    nrow(clinical_patient_clean), ncol(clinical_patient_clean), "\n")

# Check data types
cat("\n[CHECK] Column classes:\n")
print(sapply(clinical_patient_clean, class))

# Check missing values
total_na <- sum(is.na(clinical_patient_clean))
cat("\n[CHECK] Total NA values:", total_na, "\n")

na_by_col <- colSums(is.na(clinical_patient_clean))
cat("[CHECK] NA per column:\n")
print(na_by_col)

# Remove rows with missing or empty SUBTYPE
before <- nrow(clinical_patient_clean)

clinical_patient_clean <- clinical_patient_clean[
  !is.na(clinical_patient_clean$SUBTYPE) &
  clinical_patient_clean$SUBTYPE != "",
]

after <- nrow(clinical_patient_clean)

cat("\n[CLEAN] Removed rows with missing SUBTYPE.\n")
cat("Rows before:", before,
    "Rows after:", after,
    "Removed:", before - after, "\n")

# Check duplicate PATIENT_ID
dup_count <- sum(duplicated(clinical_patient_clean$PATIENT_ID))
cat("\n[CHECK] Duplicate PATIENT_ID count:", dup_count, "\n")

if (dup_count > 0) {
  clinical_patient_clean <- clinical_patient_clean[
    !duplicated(clinical_patient_clean$PATIENT_ID),
  ]
  cat("[CLEAN] Duplicate PATIENT_ID rows removed.\n")
} else {
  cat("[CLEAN] No duplicate PATIENT_ID found.\n")
}

# Executive summary
cat("\n=========================\n")
cat("EXECUTIVE SUMMARY (PATIENT DATA)\n")
cat("=========================\n")
cat("Final number of patients:", nrow(clinical_patient_clean), "\n")
cat("Total NA values:", sum(is.na(clinical_patient_clean)), "\n")
cat("Duplicate PATIENT_ID:", sum(duplicated(clinical_patient_clean$PATIENT_ID)), "\n")

# Display first 10 rows (required in assignment)
head(clinical_patient_clean, 10)

```

```{r}
# =========================
# Assignment 2 - Part c
# Quality check + clean  clinical Sample data
# =========================

# Keep only PATIENT_ID and SAMPLE_ID
clinical_sample_clean <- clinical_sample[, c("PATIENT_ID", "SAMPLE_ID")]

cat("\n[CLEAN] Dimensions after keeping PATIENT_ID and SAMPLE_ID:",
    nrow(clinical_sample_clean), ncol(clinical_sample_clean), "\n")

# Check data types
cat("\n[CHECK] Column classes:\n")
print(sapply(clinical_sample_clean, class))

# Check missing values
total_na <- sum(is.na(clinical_sample_clean))
cat("\n[CHECK] Total NA values:", total_na, "\n")

na_by_col <- colSums(is.na(clinical_sample_clean))
cat("[CHECK] NA per column:\n")
print(na_by_col)

# Remove rows with missing PATIENT_ID or SAMPLE_ID
before <- nrow(clinical_sample_clean)

clinical_sample_clean <- clinical_sample_clean[
  !is.na(clinical_sample_clean$PATIENT_ID) &
  clinical_sample_clean$PATIENT_ID != "" &
  !is.na(clinical_sample_clean$SAMPLE_ID) &
  clinical_sample_clean$SAMPLE_ID != "",
]

after <- nrow(clinical_sample_clean)

cat("\n[CLEAN] Removed rows with missing IDs.\n")
cat("Rows before:", before,
    "Rows after:", after,
    "Removed:", before - after, "\n")

# Check duplicate SAMPLE_ID (should be unique)
dup_sample <- sum(duplicated(clinical_sample_clean$SAMPLE_ID))
cat("\n[CHECK] Duplicate SAMPLE_ID count:", dup_sample, "\n")

if (dup_sample > 0) {
  clinical_sample_clean <- clinical_sample_clean[
    !duplicated(clinical_sample_clean$SAMPLE_ID),
  ]
  cat("[CLEAN] Duplicate SAMPLE_ID rows removed.\n")
} else {
  cat("[CLEAN] No duplicate SAMPLE_ID found.\n")
}

# Check duplicate PATIENT_ID (can exist — one patient may have multiple samples)
dup_patient <- sum(duplicated(clinical_sample_clean$PATIENT_ID))
cat("[CHECK] Duplicate PATIENT_ID count:", dup_patient, "\n")

# Executive summary
cat("\n=========================\n")
cat("EXECUTIVE SUMMARY (SAMPLE DATA)\n")
cat("=========================\n")
cat("Final number of samples:", nrow(clinical_sample_clean), "\n")
cat("Total NA values:", sum(is.na(clinical_sample_clean)), "\n")
cat("Duplicate SAMPLE_ID:", sum(duplicated(clinical_sample_clean$SAMPLE_ID)), "\n")

# Display first 10 rows (required)
head(clinical_sample_clean, 10)
```

The clinical patient dataset contained 514 entries, of which 7 were removed due to missing subtype information, resulting in 507 labeled patients. The clinical sample dataset contained 514 samples with no missing identifiers. The difference in counts reflects samples belonging to patients without subtype labels. After merging the datasets on PATIENT_ID, the resulting dataset contains 507 samples with valid subtype classification.

```{r}

#####################  Part D Step 1: Transpose mRNA  #########################

# Extract gene names
gene_names <- mrna_clean$Hugo_Symbol

# Extract numeric matrix
mrna_matrix <- as.matrix(mrna_clean[, -1])

# Transpose (samples become rows)
mrna_transposed <- t(mrna_matrix)

# Convert to data frame
mrna_transposed <- as.data.frame(mrna_transposed)

# Assign gene names as column names
colnames(mrna_transposed) <- gene_names

# Add SAMPLE_ID column
mrna_transposed$SAMPLE_ID <- rownames(mrna_transposed)

# Move SAMPLE_ID to first column
mrna_transposed <- mrna_transposed[, c("SAMPLE_ID", gene_names)]

# Reset rownames
rownames(mrna_transposed) <- NULL

cat("[CHECK] Dimensions after transpose:",
    nrow(mrna_transposed), ncol(mrna_transposed), "\n")
```

```{r}
#####################  Part D Step 2: Merge Sample + Patient #####################

sample_patient_df <- merge(
  clinical_sample_clean,
  clinical_patient_clean,
  by = "PATIENT_ID"
)

cat("[CHECK] Dimensions after merging sample + patient:",
    nrow(sample_patient_df), ncol(sample_patient_df), "\n")
```

```{r}
#####################  Part D Step 3: Merge with mRNA #####################################

final_df <- merge(
  sample_patient_df[, c("SAMPLE_ID", "SUBTYPE")],
  mrna_transposed,
  by = "SAMPLE_ID"
)

cat("[CHECK] Final dimensions:",
    nrow(final_df), ncol(final_df), "\n")
```

```{r}
head(final_df, 10)
```

```{r}
#####################  Final Step: Remove Zero-Variance Genes ######################

# Identify gene columns (exclude SAMPLE_ID and SUBTYPE)
gene_cols <- setdiff(colnames(final_df), c("SAMPLE_ID", "SUBTYPE"))

# Compute variance for each gene across samples
gene_var_final <- sapply(final_df[, gene_cols, drop = FALSE], var, na.rm = TRUE)

# Count zero-variance genes
zero_var_count_final <- sum(gene_var_final == 0, na.rm = TRUE)

cat("\n[CHECK] Zero-variance genes in final dataset:", zero_var_count_final, "\n")

# Remove zero-variance genes
if (zero_var_count_final > 0) {
  before_cols <- ncol(final_df)
  
  final_df <- final_df[, c("SAMPLE_ID", "SUBTYPE",
                           names(gene_var_final[gene_var_final != 0]))]
  
  after_cols <- ncol(final_df)
  
  cat("[CLEAN] Removed zero-variance genes.\n")
  cat("Columns before:", before_cols,
      "Columns after:", after_cols,
      "Removed:", before_cols - after_cols, "\n")
} else {
  cat("[CLEAN] No zero-variance genes found in final dataset.\n")
}
```

```{r}
write.csv(final_df,
          "data/LGG_cleaned_expression_dataset.csv",
          row.names = FALSE)
```




```{r}
# Task 1: Read the cancer subtype annotated mRNA expression dataset

lgg_data <- read.csv("data/LGG_cleaned_expression_dataset.csv")

# Display the first 10 rows
head(lgg_data, 10)
```


```{r} 
#Task 2: 

# Load cleaned dataset

lgg_data <- read.csv("data/LGG_cleaned_expression_dataset.csv")
```



```{r} 
#Create a summary table

#This table shows ERBB2 statistics for each subtype. 

library(dplyr)

erbb2_summary <- lgg_data %>%
  group_by(SUBTYPE) %>%
  summarise(
    Mean_ERBB2 = mean(ERBB2, na.rm = TRUE),
    Median_ERBB2 = median(ERBB2, na.rm = TRUE),
    SD_ERBB2 = sd(ERBB2, na.rm = TRUE),
    Samples = n()
  )

erbb2_summary
```




```{r}
install.packages("ggplot2")
```


```{r} 
#ggplot2 

library(ggplot2)

ggplot(lgg_data, aes(x = SUBTYPE, y = ERBB2, fill = SUBTYPE)) +
  geom_boxplot() +
  labs(
    title = "Distribution of ERBB2 Expression by LGG Subtype",
    x = "LGG Tumor Subtype",
    y = "ERBB2 Expression"
  ) +
  theme_minimal()
```
What the plot shows ,From your visualization:

 1 LGG_IDHwt

Much higher ERBB2 expression
Very large spread 
Several extreme values

 2 LGG_IDHmut-codel

Lower expression
Smaller spread

 3 LGG_IDHmut-non-codel

Similar to IDHmut-codel
Also relatively tight distribution

The boxplot shows that ERBB2 expression varies across LGG tumor subtypes. In particular, the LGG_IDHwt subtype tends to have higher ERBB2 expression and greater variability compared to the LGG_IDHmut-codel and LGG_IDHmut-non-codel subtypes. However, there is still substantial overlap between the expression ranges of the three subtypes, especially between the two IDHmut groups. This overlap suggests that ERBB2 expression alone may not clearly distinguish between tumor subtypes.

```{r} 
#### Task 4 #########
# Step 1 — Apply log₂(x + 1) transformation

erbb2_log <- log2(lgg_data$ERBB2 + 1)

# Step 2 — Create histogram

hist(erbb2_log,
     col = "skyblue",
     main = "Histogram of log2(x + 1) Transformed ERBB2 Expression",
     xlab = "log2(ERBB2 + 1)",
     breaks = 30)
 
```

The histogram of the log₂(x + 1) transformed ERBB2 expression values shows a more symmetric distribution compared to the raw expression values. The transformation compresses large values and reduces the impact of extreme outliers, which helps reduce skewness in the data. This makes the gene expression values more suitable for statistical analysis and downstream methods such as standardization and PCA.


```{r} 
 ###### Task 5 #####

# Step 1 — Apply log₂(x + 1) to gene columns

# Apply log2(x + 1) transformation to gene expression columns
lgg_log <- lgg_data

lgg_log[, 3:ncol(lgg_log)] <- log2(lgg_log[, 3:ncol(lgg_log)] + 1)

# Step 2 — Print first 10 rows

head(lgg_log, 10)


```

The log₂(x + 1) transformation was applied to all gene expression values to reduce skewness and compress large expression values. Gene expression datasets often contain highly variable values and extreme outliers, which can negatively affect statistical analyses. Applying the log transformation stabilizes the variance and makes the distribution of gene expression values more symmetric, improving the suitability of the data for downstream analyses such as standardization and PCA.

```{r} 
# check any zero variance still there or not

 sum(apply(lgg_log[,3:ncol(lgg_log)], 2, var) == 0)
```
Task 6 is about standardizing the log-transformed gene expression values before PCA.
Why we standardize
After log transformation, genes still have different scales and variances. Some genes may have larger ranges than others. If we run PCA directly, genes with larger variance dominate the principal components.

Standardization solves this by transforming each gene so that:

          𝑥- mean
  z =     _________
    
            sd



This makes every gene have:
mean = 0
standard deviation = 1

This ensures all genes contribute equally to PCA and distance calculations, which is important for high-dimensional genomic data.

```{r} 
# Task 6 is about standardizing the log-transformed gene expression values before PCA.

# Copy the log-transformed dataset
lgg_scaled <- lgg_log

# Standardize gene expression values
lgg_scaled[,3:ncol(lgg_scaled)] <- scale(lgg_log[,3:ncol(lgg_log)])

# Display first 10 rows
head(lgg_scaled, 10)

```
 This means:
 negative values → below average expression
 positive values → above average expression

The log-transformed gene expression values were standardized using z-score normalization so that each gene has mean 0 and standard deviation 1 across samples. This step ensures that genes with larger expression ranges do not dominate the analysis. Standardization is particularly important before applying PCA.


`     ################ PCA ####################

```{r} 
# Task 7 is where real dimensionality reduction begins.PCA will work poorly if we keep all genes because many genes have very little variation across samples. Those genes contain almost no useful information for distinguishing tumor subtypes.

# So the idea is to keep only genes with higher variance.
```

```{r}
#####################  Histogram with Percentile Lines #####################################

# Compute variance
gene_cols <- setdiff(colnames(final_df), c("SAMPLE_ID", "SUBTYPE"))
gene_var <- apply(final_df[, gene_cols], 2, var)

# Log scale
log_var <- log10(gene_var + 1)

# Plot histogram
hist(log_var,
     breaks = 100,
     col = "lightblue",
     border = "white",
     main = "Log-Scaled Gene Variance Distribution",
     xlab = "log10(Variance + 1)")

# Percentiles (10% to 90%)
percentiles <- quantile(log_var, probs = seq(0.1, 0.9, 0.1))

# Add vertical dotted lines
abline(v = percentiles,
       col = "red",
       lty = 2,     # dotted
       lwd = 1.5)

# Optional: label them
text(x = percentiles,
     y = par("usr")[4] * 0.9,
     labels = paste0(seq(10,90,10), "%"),
     col = "red",
     srt = 90,
     cex = 0.7,
     pos = 4)
```

A log10(variance + 1) transformation was applied to the gene variance values to reduce skewness and improve visualization
log10(variance + 1) = compressed measure of how much a gene varies


Based on the variance distribution, the bottom 30% of genes were removed using a percentile-based threshold. This corresponds approximately to a log10(variance + 1) value of 3.3, which translates to a variance of about 2000 on the original scale.

```{r}
#####################  Task 7: Variance Filtering #####################################

gene_cols <- setdiff(colnames(final_df), c("SAMPLE_ID", "SUBTYPE"))

gene_var <- apply(final_df[, gene_cols], 2, var)

# Remove bottom 30% low-variance genes
threshold <- quantile(gene_var, 0.30)

selected_genes <- names(gene_var[gene_var > threshold])

final_filtered <- final_df[, c("SAMPLE_ID", "SUBTYPE", selected_genes)]

cat("Genes before:", length(gene_var), "\n")
cat("Genes after:", length(selected_genes), "\n")
cat("Genes removed:", length(gene_var) - length(selected_genes), "\n")

head(final_filtered, 10)

```



```{r}
#####################  Standardize Filtered Data #####################################

# Copy filtered dataset
final_filtered_scaled <- final_filtered

# Identify gene columns (exclude metadata)
gene_cols <- setdiff(colnames(final_filtered), c("SAMPLE_ID", "SUBTYPE"))

# Apply z-score standardization
final_filtered_scaled[, gene_cols] <- scale(final_filtered[, gene_cols])

# Show first 10 rows
head(final_filtered_scaled, 10)

```

```{r}
#####################  Task 8: Write to CSV #####################################

write.csv(final_filtered_scaled,
          file = "data/LGG_preprocessed_data.csv",
          row.names = FALSE)

```




```{r}
#####################  Task 9: Train-Test Split #####################################

# Set seed 
set.seed(355434)  

# Total number of samples
n <- nrow(final_filtered_scaled)

# Create random indices for training (80%)
train_index <- sample(1:n, size = 0.8 * n)

# Split the data
train_data <- final_filtered_scaled[train_index, ]
test_data  <- final_filtered_scaled[-train_index, ]

# Check sizes
cat("Training samples:", nrow(train_data), "\n")
cat("Testing samples:", nrow(test_data), "\n")

```



```{r}
#####################  Task 10: PCA on Training Data #####################################

# Extract only gene expression columns for PCA
train_x <- train_data[, -(1:2)]   # removes SAMPLE_ID and SUBTYPE

# Run PCA on training data
pca_train <- prcomp(train_x, center = FALSE, scale. = FALSE)

# Create a plotting data frame using first two PCs
pca_train_df <- data.frame(
  PC1 = pca_train$x[, 1],
  PC2 = pca_train$x[, 2],
  SUBTYPE = train_data$SUBTYPE
)

# Plot PC1 vs PC2 colored by subtype
subtype_colors <- c("red", "blue", "green")
plot(pca_train_df$PC1, pca_train_df$PC2,
     col = subtype_colors[as.factor(pca_train_df$SUBTYPE)],
     pch = 19,
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA of Training Data (PC1 vs PC2)")

legend("topright",
       legend = levels(as.factor(pca_train_df$SUBTYPE)),
       col = subtype_colors,
       pch = 19,
       cex = 0.8)
```



```{r}
#Task 11: Construct a scree plot of the first 25 principal components
pca_train



#####################  Task 11: Scree Plot #####################################

# Variance explained
var_explained <- (pca_train$sdev)^2

# Proportion of variance
prop_var <- var_explained / sum(var_explained)

# First 25 PCs
prop_var_25 <- prop_var[1:25]

# Scree plot
plot(prop_var_25,
     type = "b",
     pch = 19,
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot (First 25 PCs)")



# Cumulative variance
cum_var_25 <- cumsum(prop_var_25)

plot(cum_var_25,
     type = "b",
     pch = 19,
     xlab = "Principal Component",
     ylab = "Cumulative Variance Explained",
     main = "Cumulative Variance (First 25 PCs)")
```

The cumulative variance plot shows that the first few principal components capture a substantial portion of the variability, with diminishing returns as more components are added. Approximately 20 principal components explain around 65% of the total variance. The curve begins to flatten after around 10–15 components, suggesting that additional components contribute relatively less information. Therefore, retaining approximately 15–20 principal components provides a reasonable balance between dimensionality reduction and information retention.


The gradual increase in cumulative variance and lack of a sharp elbow further indicate that the data does not lie in a low-dimensional space.

```{r}
#####################  Task 13: Extract Reduced Data #####################################

# Number of PCs to keep
k <- 20

# Extract first k PCs
train_pca_reduced <- data.frame(pca_train$x[, 1:k])

# Add labels back
train_pca_reduced$SUBTYPE <- train_data$SUBTYPE
train_pca_reduced$SAMPLE_ID <- train_data$SAMPLE_ID

# Reorder columns
train_pca_reduced <- train_pca_reduced[, c("SAMPLE_ID", "SUBTYPE", paste0("PC", 1:k))]
```


```{r}
#####################  Save CSV #####################################

write.csv(train_pca_reduced,
          file = "data/LGG_PCA_reduced_data.csv",
          row.names = FALSE)
```
A principal component (PC) is a combination of original variables that captures the most important patterns in the data.

🔥 What each PC is doing
👉 PC1
captures maximum variation in data
most important pattern
👉 PC2
second most important pattern
👉 PC3, PC4...
less and less important

PCA solves this:
Problem	                                   PCA solution
-------                                    ------------
too many variables        --->            reduces to few PCs
correlated genes          --->            combines them
noisy data	              --->            keeps main signal



```{r}
#✅ Task 1 — Read dataset + show first 10 rows

# Load required library
library(readr)

# Read the preprocessed dataset (before PCA)
lgg_data <- read_csv("data/LGG_preprocessed_data.csv")

# Convert subtype to factor
lgg_data$SUBTYPE <- as.factor(lgg_data$SUBTYPE)

# Show first 10 rows
head(lgg_data, 10)
```


```{r}
#✅ Task 2 — Train/Test split
library(readr)

# Read correct dataset
lgg_data <- read_csv("data/LGG_preprocessed_data.csv")

# Convert subtype
lgg_data$SUBTYPE <- as.factor(lgg_data$SUBTYPE)

# Set seed 
set.seed(355434)  

# 80/20 split
train_index <- sample(1:nrow(lgg_data), 0.8 * nrow(lgg_data))

train_data <- lgg_data[train_index, ]
test_data  <- lgg_data[-train_index, ]
```



```{r}
#✅ Task 3 — PCA on training data
# Remove ID and subtype
train_x <- train_data[, !(names(train_data) %in% c("SAMPLE_ID", "SUBTYPE"))]

# Apply PCA
pca_model <- prcomp(train_x, scale. = TRUE)

# Get PCA scores
train_pca <- as.data.frame(pca_model$x)

# Add subtype back
train_pca$SUBTYPE <- train_data$SUBTYPE

# Show first 10 rows
head(train_pca, 10)
```




```{r}
#✅ aTsk 4 — Apply  PCA to test data
# Prepare test predictors
test_x <- test_data[, !(names(test_data) %in% c("SAMPLE_ID", "SUBTYPE"))]

# Apply SAME PCA transformation
test_pca <- predict(pca_model, newdata = test_x)

test_pca <- as.data.frame(test_pca)
test_pca$SUBTYPE <- test_data$SUBTYPE

# Show first 10 rows
head(test_pca, 10)

```





```{r}
#Task-6
install.packages("ggplot2")
library(ggplot2)
library(dplyr)
library(scales)
train_counts <- as.data.frame(table(train_data$SUBTYPE))
colnames(train_counts) <- c("SUBTYPE", "Count")
train_counts$Dataset <- "Training"

test_counts <- as.data.frame(table(test_data$SUBTYPE))
colnames(test_counts) <- c("SUBTYPE", "Count")
test_counts$Dataset <- "Test"

combined_counts <- bind_rows(train_counts, test_counts) %>%
  group_by(Dataset) %>%
  mutate(Percent = Count / sum(Count),
         Label = paste0(Count, "\n(", percent(Percent, accuracy = 0.1), ")"))

ggplot(combined_counts, aes(x = SUBTYPE, y = Count, fill = Dataset)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_text(aes(label = Label),
            position = position_dodge(width = 0.75),
            vjust = -0.35,
            size = 3.7,
            lineheight = 0.95) +
  labs(
    title = "LGG Molecular Subtype Distribution",
    subtitle = "Comparison of training and test datasets",
    x = "Tumor Subtype",
    y = "Number of Samples"
  ) +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 2)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))

```

👉  data is unbalanced 

IDHmut-non-codel → largest
IDHmut-codel → medium
IDHwt → smallest

If ignore this, model will:
👉 favor the majority class
👉 give misleading accuracy



```{r}
#task -7
library(nnet)

# Keep only first 20 principal components + response
train_pca_20 <- train_pca[, c("SUBTYPE", paste0("PC", 1:20))]
test_pca_20  <- test_pca[,  c("SUBTYPE", paste0("PC", 1:20))]

# Make sure response is factor
train_pca_20$SUBTYPE <- as.factor(train_pca_20$SUBTYPE)
test_pca_20$SUBTYPE  <- as.factor(test_pca_20$SUBTYPE)

# Compute inverse-frequency class weights
class_counts <- table(train_pca_20$SUBTYPE)
class_weights <- 1 / class_counts
obs_weights <- as.numeric(class_weights[as.character(train_pca_20$SUBTYPE)])

# Fit weighted multinomial logistic regression
multi_model <- multinom(SUBTYPE ~ ., data = train_pca_20, weights = obs_weights, trace = FALSE)
```

The LGG subtype distribution is imbalanced, with some classes having significantly more samples than others. To address this, inverse-frequency class weights were incorporated into a multinomial logistic regression model so that minority classes receive greater importance during training. Additionally, only the first 20 principal components were used to reduce model complexity and prevent overparameterization.

Making sure minority subtypes are not ignored
----------------------------------------------------------------
⚙️ What is the model doing internally?

👉 It learns patterns like:

“If PC1 high + PC2 low → likely subtype A”
“If PC3 very negative → subtype B”

Basically:
👉 It finds relationships between PCs and subtype
------------------------------------------------------------------

🔥 Why we use this model (simple answer)

We use multinomial logistic regression because:

1)We need to predict multiple categories (3 subtypes)
2)It works well with numeric inputs (PCs)
3)It is simple, interpretable, and a good baseline
--------------------------------------------------------------------

“One in Ten Rule” means
for classification, people often phrase it as:

You want about 10 cases per predictor, and for categorical outcomes it is safer to think about the smallest class being the limiting factor.

Apply it to your data

From your subtype counts in the training data:

LGG_IDHmut-codel = 143
LGG_IDHmut-non-codel = 192
LGG_IDHwt = 70

The smallest training class is 70.

Using the One in Ten Rule, a reasonable upper bound is:

70 / 10 = 7
So a safe recommendation is to use about 7 principal components.

Why not 20 PCs?

Because Task 7 already exposed the problem: when you tried to fit the multinomial model with too many predictors, the model became too complex. That is exactly the kind of issue the One in Ten Rule is trying to prevent.

Therefore, an appropriate choice is to use the first 7 principal components in the multinomial logistic regression model.

```{r}

#with first 7 PCs
library(nnet)

# Keep only first 7 principal components + response
train_pca_7 <- train_pca[, c("SUBTYPE", paste0("PC", 1:7))]
test_pca_7  <- test_pca[,  c("SUBTYPE", paste0("PC", 1:7))]

# Make sure response is factor
train_pca_7$SUBTYPE <- as.factor(train_pca_7$SUBTYPE)
test_pca_7$SUBTYPE  <- as.factor(test_pca_7$SUBTYPE)

# Compute inverse-frequency class weights
class_counts <- table(train_pca_7$SUBTYPE)
class_weights <- 1 / class_counts

obs_weights <- as.numeric(class_weights[as.character(train_pca_7$SUBTYPE)])

# Fit weighted multinomial logistic regression
multi_model <- multinom(SUBTYPE ~ ., 
                        data = train_pca_7, 
                        weights = obs_weights, 
                        trace = FALSE)
```
now
What this task is :

1.fit a multinomial logistic regression model
2.use SUBTYPE as the target
3.use the first X PCs as predictors
     -in our case, X = 7
4.choose a reference level
5.justify that choice
6.print summary(model)

---------------------------------------
First: what is a reference level?

Because we have 3 subtypes, multinomial logistic regression does not estimate all 3 directly in isolation.

Instead, it picks one subtype as the baseline/reference, and compares the other two against it.

So if reference = LGG_IDHmut-non-codel, the model estimates:

LGG_IDHmut-codel vs LGG_IDHmut-non-codel
LGG_IDHwt vs LGG_IDHmut-non-codel
------------------------------------------
Justification of Which subtype should be the reference?
Best choice:  LGG_IDHmut-non-codel

Why?

Because it is the largest class in training data.

it gives a more stable baseline
comparisons are made against the most common subtype
coefficient estimates are often more stable than using the smallest class


```{r}
#summary(model)
library(nnet)

# Keep only first 7 principal components + response
train_pca_7 <- train_pca[, c("SUBTYPE", paste0("PC", 1:7))]
test_pca_7  <- test_pca[,  c("SUBTYPE", paste0("PC", 1:7))]

# Make SUBTYPE a factor
train_pca_7$SUBTYPE <- as.factor(train_pca_7$SUBTYPE)
test_pca_7$SUBTYPE  <- as.factor(test_pca_7$SUBTYPE)

# Set the reference level to the largest class
train_pca_7$SUBTYPE <- relevel(train_pca_7$SUBTYPE, ref = "LGG_IDHmut-non-codel")
test_pca_7$SUBTYPE  <- relevel(test_pca_7$SUBTYPE, ref = "LGG_IDHmut-non-codel")

# Compute inverse-frequency class weights
class_counts <- table(train_pca_7$SUBTYPE)
class_weights <- 1 / class_counts
obs_weights <- as.numeric(class_weights[as.character(train_pca_7$SUBTYPE)])

# Fit weighted multinomial logistic regression model
multi_model <- multinom(SUBTYPE ~ ., 
                        data = train_pca_7, 
                        weights = obs_weights, 
                        trace = FALSE)

# Print model summary
summary(multi_model)

```
-------
output is basically telling
how each PC affects the odds of being one subtype versus the reference subtype
-------


************** Part C  validating the classifier **********************

 give this  model new data (test set) and check if its predictions are correct.
 
 
```{r}
#✅ Step 1: Predictions on test data
predictions <- predict(multi_model, newdata = test_pca_7)
#✅ Step 2: Confusion Matrix
conf_matrix <- table(Predicted = predictions, Actual = test_pca_7$SUBTYPE)
conf_matrix
#✅ Step 3: Misclassification Rate
misclassification_rate <- 1 - sum(diag(conf_matrix)) / sum(conf_matrix)
misclassification_rate


library(ggplot2)

cm_df <- as.data.frame(conf_matrix)

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion Matrix",
    x = "Actual Subtype",
    y = "Predicted Subtype"
  ) +
  theme_minimal()
```

The confusion matrix was computed to evaluate the performance of the multinomial logistic regression model on the test data.

Misclassification rate measures how often the model makes incorrect predictions.


```{r}
# Accuracy
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)

# Misclassification rate
misclassification_rate <- 1 - accuracy

# Convert to percentage
accuracy_percent <- round(accuracy * 100, 2)
misclassification_percent <- round(misclassification_rate * 100, 2)

# Print results
cat("Accuracy:", accuracy_percent, "%\n")
cat("Misclassification Rate:", misclassification_percent, "%\n")
```


```{r}
# with 2 pcs
library(nnet)

# Keep only PC1 and PC2
train_pca_2 <- train_pca[, c("SUBTYPE", "PC1", "PC2")]
test_pca_2  <- test_pca[,  c("SUBTYPE", "PC1", "PC2")]

# Convert to factor
train_pca_2$SUBTYPE <- as.factor(train_pca_2$SUBTYPE)
test_pca_2$SUBTYPE  <- as.factor(test_pca_2$SUBTYPE)

# Set reference level (same as before)
train_pca_2$SUBTYPE <- relevel(train_pca_2$SUBTYPE, ref = "LGG_IDHmut-non-codel")
test_pca_2$SUBTYPE  <- relevel(test_pca_2$SUBTYPE, ref = "LGG_IDHmut-non-codel")

# Fit model (you can keep weights OR skip — both acceptable)
model_pc2 <- multinom(SUBTYPE ~ ., data = train_pca_2, trace = FALSE)

# Predictions
pred_pc2 <- predict(model_pc2, newdata = test_pca_2)

# Confusion matrix
cm_pc2 <- table(Predicted = pred_pc2, Actual = test_pca_2$SUBTYPE)

# Misclassification rate
mis_pc2 <- 1 - sum(diag(cm_pc2)) / sum(cm_pc2)

# Convert to percentage
mis_pc2_percent <- round(mis_pc2 * 100, 2)

cat("Misclassification Rate (PC1 & PC2):", mis_pc2_percent, "%\n")
```
A multinomial logistic regression model was fitted using only the first two principal components as predictors. The misclassification rate on the test data increased compared to the model using seven principal components. This indicates that using only two principal components does not capture sufficient variation in the data to accurately distinguish between LGG subtypes, leading to reduced classification performance.

👉 More PCs ≠ always better
👉 Too few PCs = underfitting

```{r}
#lets try 100 pcs

#Step 1: Use first 100 PCs
train_pca_100 <- train_pca[, c("SUBTYPE", paste0("PC", 1:100))]
test_pca_100  <- test_pca[,  c("SUBTYPE", paste0("PC", 1:100))]

#Step 2: Factor + reference

train_pca_100$SUBTYPE <- as.factor(train_pca_100$SUBTYPE)
test_pca_100$SUBTYPE  <- as.factor(test_pca_100$SUBTYPE)

train_pca_100$SUBTYPE <- relevel(train_pca_100$SUBTYPE, ref = "LGG_IDHmut-non-codel")
test_pca_100$SUBTYPE  <- relevel(test_pca_100$SUBTYPE, ref = "LGG_IDHmut-non-codel")

#Step 3: Fit model

library(nnet)

model_pc100 <- multinom(SUBTYPE ~ ., 
                        data = train_pca_100, 
                        trace = FALSE,
                        MaxNWts = 10000)

#Step 4: Predictions

pred_pc100 <- predict(model_pc100, newdata = test_pca_100)

#Step 5: Confusion matrix + error

cm_pc100 <- table(Predicted = pred_pc100, Actual = test_pca_100$SUBTYPE)

mis_pc100 <- 1 - sum(diag(cm_pc100)) / sum(cm_pc100)

mis_pc100_percent <- round(mis_pc100 * 100, 2)

cat("Misclassification Rate (PC1–PC100):", mis_pc100_percent, "%\n")
```
The misclassification rate did not significantly improve compared to the model using 7 principal components, indicating that additional components do not contribute meaningful predictive information and may introduce noise.

```{r}
#✅ Step 1: Create comparison table
comparison_df <- data.frame(
  Model = c("PC1–PC2", "PC1–PC7", "PC1–PC100"),
  Misclassification = c(mis_pc2, misclassification_rate, mis_pc100)
)

# Convert to percentage
comparison_df$Misclassification_Percent <- round(comparison_df$Misclassification * 100, 2)

#✅ Step 2: Clean bar chart

library(ggplot2)

ggplot(comparison_df, aes(x = Model, y = Misclassification_Percent, fill = Model)) +
  geom_col(width = 0.6) +
  
  geom_text(aes(label = paste0(Misclassification_Percent, "%")),
            vjust = -0.4,
            size = 5,
            fontface = "bold") +
  
  labs(
    title = "Comparison of Misclassification Rates",
    subtitle = "Effect of Number of Principal Components",
    x = "Model (Number of PCs Used)",
    y = "Misclassification Rate (%)"
  ) +
  
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  
  theme_minimal() +
  theme(legend.position = "none")

```
👉 The model using the first 7 principal components is the best, as it achieves the lowest misclassification rate and provides the most reliable predictions.

--------------------------------------------------------

What is baseline accuracy?

From test data:

LGG_IDHmut-non-codel = 56
total = 102

👉 Baseline accuracy ≈

56 / 102 ≈ 55%

So what is success?

👉 model should beat ~55% accuracy

Accuracy ≈ 96%

👉 That is:

far above baseline ✅
very strong model ✅

-------------------------------------------------

analytical question

How does the number of principal components (PCs) affect the performance of a multinomial logistic regression model in classifying LGG tumor subtypes compared to a baseline classifier?


